---
title: "Design Motivation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design Motivation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The motivating idea behind this package is that every data point has context,
and an object storing data points should provide access to this context. The
`data.frame` object enables this to some extent by allowing us to annotate
data points with labels (the `"row.names"`), but it is not flexible
enough to handle cases where the context involves more than one attribute. We
fix this by replacing the `"row.names"` attribute of a `data.frame` with a new
attribute, `"keys"`, that stores potentially multi-component labels
(*composite keys*) that identify each row in the data set.


This package takes the view that a *variable* is a map (function) between two
sets, called the *frame* (domain) and the *type* (domain) of the variable. We
refer to members of the *frame* as *keys* and members of the *type* as
*values*. A *data point* is a (variable, key, value) triple. This differs from
the "tidy" view of data that considers a data point to be a (variable, value)
pair. We believe that the context of the value is important and helpful,
especially for linking downstream analyses to the original data sources.

```{r}
library(frame)
```

## Keys

### Requirements

We allow two types of keys: simple and composite. A simple key is one of the
following basic R types: logical, integer, double, complex, or character. For
character keys, we require that the value be encoded in valid UTF-8.  A
composite string is a tuple of simple keys, each of which is called a
component or attribute of the key.

Beyond the type constraints, we require that each key (or composite key in the
case of multi-attribute keys) correspond to a single data point. To this end,
we require that the keys (or composite keys) be unique values.  We do not
require that keys be unique after Unicode normalization, and we allow the
empty string (`""`) as a key.


### Component names

An `m`-component key can optionally have names for the attributes. For
example, if we frame a set of hourly measurements by a two-component key,
we can optionally name these components, say `"day"` and `"hour"`. It is not
necessary that we give the components names. Making component names optional
allows us to seamlessly convert between, say, a matrix with row and column
names, and a framed data set with a two-component key corresponding to the row
and column indices. Allowing anonymous keys also allows us to convert
from a `data.frame` with `row.names` to a keyed data set with an anonymous
key (taken from the `row.names`).


## Data sets

We provide the `dataset` type as an extension of `data.frame` that allows
specifying keys for the rows. We construct a dataset as in the following
example:

```{r}
(x <- dataset(a = letters[10:14], b = rnorm(5), c = 3 * -2:2))
```

Columns can be vector-like or matrix-like (having one or two dimensions). This
allows for the possibility of having nested datasets within datasets:

```{r}
(y <- dataset(nest = x, simple = runif(5)))
```


## Indexing

### Logical or numeric subscript

For dataset `x` and logical or numeric vector `i`, the indexing operations
`x[i,]` and `x[i,j]` behave like the counterpart `data.frame` operations,
except that `drop = FALSE` is the default. We apply the same row subsetting to
the keys, with the following exception. If `keys(x)` is non-`NULL` and `i` is
numeric with duplicate entries, then we append a column to the keys to ensure
that `keys(x[i,])` has unique rows. The added column is the appearance number
of each row in `x[i,]`. We demonstrate this behavior below:

```{r}
x <- dataset(a = 1:26)
keys(x) <- letters
x[c(1, 2, 1, 1, 3, 2), ]
```

### Key subscript

(more to come...)
