---
title: "Indexing Operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Indexing Operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(frame)
options(frame.rows = 4)
options(frame.wrap = 0)
```

The indexing and operations supported by *frame* `dataset` objects are similar
to those supported by `data.frame` objects, but there are a few extensions and
differences.

## Sample Data

We will illustrate the indexing operations with a sample dataset with
two key columns, and 5 data columns, one of which is a matrix.

```{r}
x <- cbind.dataset(name = rownames(mtcars),
                   mtcars[1:3],
                   dataset(group = mtcars[4:5]))
keys(x) <- dataset(k1 = rep(LETTERS[1:4], 8),
                   k2 = rep(1:8, each = 4))
print(x)
```

Note the dimensions of `x`:

```{r}
length(x)
dim(x)
```

The matrix column `"group"` gets counted as a single column in `x`.


## Extracting a column

We can extract a column using `$` or `[[`.

```{r}
x[[1]]
x[["cyl"]]
x$cyl
```

We can also using `[` with `drop = TRUE`.

```{r}
x[, 1, drop = TRUE]
x[, "cyl", drop = TRUE]
```

Unlike `data.frame` objects, we do not allow partial name matching.

```{r, error = TRUE}
x$cy
```

Another difference from `data.frame` objects is that `drop = FALSE` is
the default, so that `x[, 1]` returns a `dataset`.


## Selecting columns

We can select a subset of columns using `[`. This works in three modes:
by name, by index, and by logical mask.

### By name

```{r}
x[, "name"]
x[, c("mpg", "disp")]
x[, c("mpg", "group", "mpg")]
```

Note that we allow duplicate names.

### By index

```{r}
x[, 1]
x[, c(2, 4)]
x[, c(2, 5, 2)]
```

### By logical mask

```{r}
x[, c(TRUE, FALSE, FALSE, FALSE, FALSE)]
x[, c(FALSE, TRUE, FALSE, TRUE, FALSE)]
```


## Selecting rows

We can select a subset of rows using `[` by key, by index, and by logical
mask; for `dataset` objects with `rownames`, we can also select by name.

### By key

### By index

### By logical mask

### By name

Unlike `data.frame` objects, not all `dataset` objects have row names. Our
example dataset `x` is one such example with none:

```{r}
rownames(x)
```

Only datasets with a single character key have row names; in these cases the
keys are the rownames.

```{r}
(y <- as_dataset(mtcars))
rownames(y)
```

For datasets with row names, we can select rows using these names.
```{r}
y[c("Duster 360", "Mazda RX4 Wag", "Lotus Europa"), ]
```


## Extracting a row

We can extract a single row by selecting a subset of size one and specifying
`drop = TRUE`. This returns a list.

```{r}
x[1, , drop = TRUE]
```
